Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-04-30 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
add regular rhat in figure 3, adjust methods section to include hypotheses and better structure with qs and thetas.

@@

\caption{Typical convergence versus pathological non-convergence.}\label{fig:non-conv}
\end{figure}

@@

@@

\caption{Convergence diagnostics pathological non-convergence versus typical convergence.}\label{fig:diagnostics}
\end{figure}

@@

\textbf{Add to this section: 1) explicit mention of simulation
conditions; 2) emphasize that the plots are averages across repetitions,
not within MICE; 3) sample effects due to single complete dataset}

The aim of the simulation study is to evaluate the impact of inducing
non-convergence on several quantities of scientific interest \(Q\), and
to subsequently evaluate the performance of \(\widehat{R}\) and \(AC\)
in identifying the non-convergence.

Non-convergence will be induced by: 1) varying the missingness
proportions in \(y\) (\(p_{mis} =.05,.25,.50,.75,.95\)), and 2)
terminating the MICE algorithm at different imputation chain lengths
(\(t = 1, 2,..., 100\)). The assumption underlying the different number
of iterations is that the algorithm generally does reach convergence at
\(t=1\). The initial values in the first iteration are sampled randomly
from the set of observed datapoints. \textbf{As the number of iterations
goes up, the imputation chains become independent of the initial values
until the point at which an extra added iteration does not lead to a
more converged state.} The second set of experimental conditions under
consideration--the missingness proportions--are chosen to reflect the
difficulty of the missingness problem. The inherent assumption is that
low missingness proportions will lead to quick algorithmic convergence,
since there is a lot of information in the observed data. Higher
missingness proportions then cause slower convergence. However, if the
fraction of missing information is very high, there is so little
information in the data that the random component in the algorithm will
take the overhand and a stable but very uncertain (high variance) state
will be reached.

\hypertarget{hypothesis}{%
\subsection{Hypothesis}\label{hypothesis}}

We hypothesize that simulation conditions with a higer \(p_miss\) and
lower \(t\) have inferior convergence. We assume that when \(\bar{Q}\)
is an unbiased and confidence valid estimate of \(Q\), the algorithm is
\emph{sufficiently} converged.

\textbf{(bruggetje hier)} For multiple imputation algorithms, it holds
that convergence is reached when there is no dependency between
subsequent iterations of imputation chains (\(AC = 0\)), and chains
intermingle such that the only difference between the chains is caused
by the randomness induced by the algorithm (\(\widehat{R} = 1\)).
\textbf{We expect that a completely converged state will not be
reached.}

Based on an empirical finding \citep{lace07}, we hypothesize that
\(\widehat{R}\) will over-estimate non-convergence of MI algorithms
\textbf{explain why?}. Lacerda concluded that rhat dropped below 1.2
after at least 50 iterations. Therefore we hypothesize that Vehtari's
proposed threshold of \(\widehat{R} < 1.01\) will be too stringent for
diagnosing convergence in iterative imputation. This over-estimation
may, however, be diminished because \(\widehat{R}\) can falsely diagnose
convergence if initial values of the algorithm are not appropriately
over-dispersed \citep[p.~437]{broo98}. In e.g.~\texttt{mice}, initial
values are chosen randomly from the observed data. Therefore, we cannot
be certain that the initial values are over-dispersed. We expect this to
have little effect on the hypothesized performance of \(\widehat{R}\).
\textbf{(Add actual hypothesis: we thought that mice would converge
sooner than \(\widehat{R}\) would indicate that it did)} We hypothesize
that high AC values are implausible in converged MI algorithms.
\textbf{Added later: High \(AC\)-values are implausible in MI
procedures. That is, the randomness induced by the MI algorithm
effectively mitigates the risk of dependency within chains. But this
doesn't seem the case in practice??}

@@

\begin{verbatim}
# pseudo-code of simulation 
1. simulate data 
for (number of simulation runs from 1 to 1000)
 for (missingness proportions 5%, 25%, 50%, 75% and 95%)
  2. create missingness
  for (number of iterations from 1 to 100)
   3. impute missingness
   4. perform analysis of scientific interest
   5. compute convergence diagnostics 
   6. pool results across imputations
   7. compute performance measures
 8. combine outcomes of all missingness proportions
9. aggregate outcomes across simulation runs 
\end{verbatim}

The data-generating mechanism is a multivariate normal distribution,
representing person-data on three predictor variables (from an
unspecified social scientific field of study). Let our predictor space
be defined as \begin{align*}
\begin{pmatrix}X_1\\
X_2\\
X_3
\end{pmatrix} \sim \mathcal{N}
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0
\end{pmatrix}
\end{bmatrix}\!\!\text{.}\\[2\jot]
\end{align*}

In this study, sampling variance is not of interest. Therefore, a single
complete set may serve as comparative truth in all simulation runs
\citep{vink14}. A finite population of \(N=1000\) is simulated using the
\texttt{mvtnorm} package \citep{mvtnorm}. Subsequently, a fourth
variable is constructed as outcome variable in a multiple linear
regression problem. Let \[
Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
\] where \(i = 1, 2,..., N\) and \(\epsilon \sim \mathcal{N}(0, 100)\).

The complete data is \emph{amputed} once for each simulation repetition
with function \texttt{mice::ampute()}. The missingness is multivariate,
and the probability to be missing is the same for all \(n \times k\)
cells in \(y\) \citep[`MCAR';][]{rubin87}. Proper performance of the
convergence diagnostics under at least `missing completely at random'
missingness mechanism is necessary to demonstrate the appropriateness of
\(\widehat{R}\) and \(AC\) as convergence diagnostics. However, results
may not be extrapolated to other missingness mechanisms. The missinness
proportion varies between conditions (\(p_{mis} =.05,.25,.5,.75,.95\)).

Missing datapoints in \(y\) are imputed with the \texttt{mice} package
\citep{mice}. All MI procedures are performed with Bayesian linear
regression imputation (\texttt{method\ =\ "norm"}), and five imputation
chains (\texttt{m\ =\ 5}). The number of iterations varies between
simulation conditions (\texttt{maxit\ =} \(1, 2, \dots, 100\)).

The methods under evaluation, \(\widehat{R}\) and \(AC\), are computed
for four scalar summaries \(\theta\): chain means (i.e., the mean in
each \(y_{imp, m, j}\)), chain variances (i.e., the variance in each
\(y_{imp, m, j}\)), the statistic of scientific interest \(\hat{Q}\) in
each \{\(y_{obs}, y_{imp, m}\)\}, and the first eigenvalue of the
variance-covariance matrix in each \{\(y_{obs}, y_{imp, m}\)\},
\(\lambda_1\). We calculate \(\widehat{R}\) by implementing Vehtari et
al.'s \citeyearpar{veht19} recommendations, and \(AC\) as the
correlation between the \(t^{th}\) and the \((t+1)^{th}\) iteration. For
each \(\theta\) we estimate the corresponding scientific estimand \(Q\).

The performance of two methods on each of these scalar summaries will be
evaluated with several estimands, i.e.~quantities of scientific interest
\(Q\). For each \(\theta\), one estimand \(Q\) is defined that may be of
interest in empirical research. The \(\widehat{R}\) and \(AC\) values
with chain means as \(\theta\)s are evaluated against the bias in
univariate mean estimates. Similarly, the performance measure for
\(\widehat{R}\) and \(AC\) applied to chain variances is the bias in
estimated standard deviations. To evaluate the performance of
\(\widehat{R}\) and \(AC\) on the eigenvalues, we will use bias in the
coefficient of determination, \(R^2\). Additionally, we will evaluate
the estimated regression coefficients, the coverage rate of the CI95\%
of the regression estimates, and the CI length (\textbf{see definitions
in old Methods}). \textbf{remove the term performance measure here and
add an extra paragraph to define performance measures as bias in all
estimands, and coverage rate and CI length of regression coefficients.}

We consider four quantities of scientific interest, namely the
descriptive statistics of all variables (mean and standard deviation),
\(\beta_2\), and the percentage of variance in \(Y\) explained
(coefficient of determination \(R^2 \times 100\)). We solve a multiple
linear regression problem, where outcome \(Y\) is regressed on
predictors \(X_1\), \(X_2\) and \(X_3\)

\[Y \sim \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3.\]

@@

The performance of \(\widehat{R}\) and \(AC\) is assessed by comparing
\(\bar{Q}\)s with \(Q\)s. For each \(Q\), we compute bias as
\(\bar{Q} - Q\). For \(Q=\beta_j\), we also compute the empirical
coverage rate (CR). CR is defined as the percentage of simulation
repetitions in which the 95\% confidence interval (CI) around
\(\bar{Q}\) covers the true estimand \(Q\). Let

\[\text{CI} = \bar{Q} \pm t_{(M-1)} \times SE_{\bar{Q}},\]

where \(t_{(M-1)}\) is the quantile of a \(t\)-distribution with \(M-1\)
degrees of freedom, and \(SE_{\bar{Q}}\) is the square root of the
pooled variance estimate. \textbf{Remove this? } Under-estimating the
variance of \(\bar{Q}\) may yield spurious inferences. Confidence
interval width (CIW) is defined as the difference between the lower and
upper bound of the 95\% confidence interval (CI) around \(\bar{Q}\) From
bias and CIW, we calculate empirical coverage rates. Coverage rate is
the proportion of simulations in which \(Q\) is between the bounds of
the CI95\% around \(\bar{Q}\).

@@

@@

>>>>

The goal of imputation through e.g.~FCS/MICE is to pbtain a multivariate
distribution by iterating over the sequence of univariate imputations.
Traditionally, the inspection of the multivariate convergence is done by
visually inspecting the univariate chain means and variances over the
iterations, or by tracking the Rhat and autocorrelation. The downside to
these approaches is that they either focus on the univariate state
space, or primarily track the change over the iterations of a
multivariate outcome conform the scientific model of interest. Focusing
on the convergence of these outcome parameters may influence this
procedure in the sense that the model of evaluation favors the model of
interest. Ideally, one would like to evaluate a model-independent
parameter that summarizes the multivariate nature of the data. We
propose eigenvalue as such a parameter, because it\ldots.

@@

\caption{Typical convergence versus pathological non-convergence. Please note that this is the same data with a different imputation model, leading to different imputations (see the y axis!). Wel min of meer dezelfde origin op iteration 1.}\label{fig:non-conv}
\end{figure}

@@

We assume that complete convergence, defined as \(\widehat{R} = 1\) and
\(AC = 0\), will not be observed. Because even in the most converged
state, the algorithm will show some signs of non-mixing and
non-stationarity. Upon \emph{sufficient} convergence, the imputation
chains will intermingle such that the only difference between the chains
is caused by the randomness induced by the algorithm
(\(\widehat{R} > 1\)), and there may be some dependency between
subsequent iterations of imputation chains (\(AC > 0\)).

@@

\caption{Convergence diagnostics applied on the imputation algorithms of Figure 2.}\label{fig:diagnostics}
\end{figure}

@@

The aim of the simulation study is to evaluate the impact of inducing
non-convergence in the MICE algorithm on several quantities of
scientific interest \(Q\). Subsequently we will evaluate how well
\(\widehat{R}\) and \(AC\) perform in identifying the effects of
non-convergence. And finally, we will formulate an informed advice on
the requirements to safely assume \emph{sufficient} convergence in
practice. That is, we assume that the algorithm is \emph{sufficiently}
converged when each estimate \(\bar{Q}\) is an unbiased and confidence
valid estimate of the corresponding estimand \(Q\).

Non-convergence will be induced by: 1) increasing the missingness
proportion \(p_{mis}\) in dataset \(y\) incrementally, and 2)
terminating the imputation algorithm at a varying number of iterations
\(T\). The first set of simulation conditions---the missingness
proportions---is chosen to reflect the difficulty of the missingness
problem. The underlying assumption is that low missingness proportions
lead to quick algorithmic convergence, since there is a lot of
information in the observed data. Higher missingness proportions should
yield slower convergence. Unless, however, the fraction of missing
information is so high that the random component in the imputation
algorithm outweighs the information in the observed data. Then,
convergence to a stable but highly variable state may be reached
instanteneously. We assume that the incremental missingness proportions
in our study will result in a corresponding increase in signs of
non-convergence.

The assumption inherent to the second set of simulation conditions---the
number of iterations---is that terminating the imputation algorithm too
early causes non-convergence. Generally, the algorithm will not reach
convergence if \(T=1\), because the imputed values in the first
iteration (at \(t=1\)) depend on the initial values of the algorithm
(which are sampled randomly from the set of observed datapoints). As the
number of iterations increases, the imputation chains should become
independent of the initial values, until the point at which adding an
extra iteration does not lead to a more converged state. We assume that
we can induce non-convergence at least in conditions where \(T\) is
smaller than the default number of iterations in \texttt{mice}, \(T=5\).

\hypertarget{hypotheses}{%
\subsection{Hypotheses}\label{hypotheses}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  We expect that simulation conditions with a high missingness
  proportion \(p_{mis}\) and a low number of iterations \(T\) will
  result in biased, invalid estimates of the quantities of scientific
  interest, \(Q\)s.
\item
  We hypothesize that \(\widehat{R}\) and \(AC\) will correctly identify
  signs of non-convergence in those simulation conditions where the
  \(\bar{Q}\)s are \emph{not} unbiased and confidence valid estimates of
  the \(Q\)s.
\item
  We hypothesize that the recommended thresholds to diagnose
  non-convergence with \(\widehat{R}\) (\(\widehat{R} > 1.2\),
  \(\widehat{R} > 1.1\), and \(\widehat{R} > 1.01\)) may be too
  stringent for iterative imputation applications. In an empirical
  study, where \(\widehat{R}\) was used to inform the required
  imputation chain length, it took as many as 50 iterations to overcome
  the conventional non-convergence threshold \(\widehat{R}>1.2\). Yet,
  scientific estimates were insensitive to continued iteration after
  \(t>5\) \citep{lace07}. We therefore suspect that \(\widehat{R}\) may
  over-estimate signs of non-convergence in iterative imputation
  algorithms. In contrast to this, it may occur that the signs of
  non-convergence are under-estimated by \(\widehat{R}\), in exceptional
  cases where the initial values of the algorithm are not appropriately
  over-dispersed \citep[p.~437]{broo98}. In e.g.~\texttt{mice}, initial
  values are chosen randomly from the observed data, hence we cannot be
  certain of over-dispersion in the initial values. In practice, we do
  not expect this to cause problems for identifying non-convergence with
  \(\widehat{R}\).
\item
  We expect that high \(AC\) values are implausible in iterative
  imputation algorithms with typical convergence. That is, after only a
  few iterations, the randomness induced by the algorithm will
  effectively mitigate the risk of dependency within chains.
\end{enumerate}

@@

\begin{verbatim}
# pseudo-code of simulation 
1. simulate data 
for (number of simulation runs from 1 to 1000)
 for (missingness proportions 5%, 25%, 50%, 75% and 95%)
  2. create missingness
  for (number of iterations from 1 to 100)
   3. impute missingness
   4. perform analysis of scientific interest
   5. compute non-convergence diagnostics 
   6. pool results across imputations
   7. compute performance measures
 8. combine outcomes of all missingness proportions
9. aggregate outcomes across simulation runs 
\end{verbatim}

\hypertarget{data-generating-mechanism.}{%
\subsubsection{Data-generating
mechanism.}\label{data-generating-mechanism.}}

In this study, sampling variance is not of interest. Therefore, a single
complete dataset may serve as comparative truth in all simulation
repetitions \citep{vink14}. The data-generating mechanism is a
multivariate normal distribution, representing person-data on three
predictor variables in a multiple linear regression problem
\textbf{(from an unspecified social scientific field of study)}. Let the
predictor space be defined as \[
\begin{aligned}
\begin{pmatrix}X_1\\
X_2\\
X_3
\end{pmatrix} \sim \mathcal{N}
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0
\end{pmatrix}
\end{bmatrix}\!\!\text{.}\\[2\jot]
\end{aligned}
\] A finite population of \(N=1000\) is simulated using the
\texttt{mvtnorm} package \citep{mvtnorm}. Subsequently, a fourth
variable is constructed as outcome \(Y\). For each unit
\(i = 1, 2,..., N\), let \[
Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
\] where \(\epsilon \sim \mathcal{N}(0, 100)\). From the complete set,
we obtain the true values of each quantity of scientific interest \(Q\).

\hypertarget{scientific-estimands.}{%
\subsubsection{Scientific estimands.}\label{scientific-estimands.}}

We consider four types of \(Q\)s that are often of interest in empirical
research. Namely, two descriptive statistics: the mean \(\mu_j\) and
standard deviation \(\sigma_j\) of each variable
\(j = Y, X_1, X_2, X_3\). We also consider the regression coefficients
of the predictors: \(\beta_1\), \(\beta_2\), and \(\beta_3\) in
regression equation
\[Y' = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3,\] where \(Y'\)
is the expected value of the outcome. And finally, the proportion of
variance explained by the set of predictors: coefficient of
determination \(r^2 \times 100\) \textbf{(note that lower case \(r\) is
used to avoid confusion with non-convergence diagnostic
\(\widehat{R}\))}. The true values of these scientific estimands are
calculated with the R functions \texttt{base::mean()},
\texttt{stats::sd()}, and \texttt{stats::lm()} \citep{R}. Each of these
\(Q\)s is estimated with a corresponding \(\bar{Q}\), after the complete
dataset is first \emph{amputed} (with a varying missingness proportion
\(p_{mis}\)) and subsequently imputed (with a varying number of
iterations \(T\)).

\hypertarget{methods.}{%
\subsubsection{Methods.}\label{methods.}}

The complete dataset is amputed with the \texttt{mice} package
\citep[function \texttt{mice::ampute()};][]{mice}. All missingness is
multivariate, and conform a `missing completely at random' missingness
mechanism \citep{rubin87}. I.e., the probability to be missing is the
same for all \(N \times k\) cells in \(y\), conditional on the
missingness proportion (\(p_{mis} =.05,.25,.5,.75,.95\)).

Missing datapoints in \(y\) are imputed with the \texttt{mice()}
function \citep{mice}. All imputation procedures are performed with
Bayesian linear regression imputation, and five imputation chains
(\(M=5\)). The number of iterations varies between simulation conditions
(\(T=1,2,...,100\)).

After each run of the \texttt{mice} algorithm, the function
\texttt{mice::complete()} is used to combine the imputed data
\(y_{imp,m}\) with the observed data \(y_{obs}\), resulting in five
completed datasets \{\(y_{obs}, y_{imp, m}\)\}, where
\(m = 1, 2, ..., 5\).

\begin{itemize}
\item
  on the 5 imputations \(y_{imp,m}\): get mean and var within imp chains
  as theta to compute conv diag --\textgreater{} then complete with
  \texttt{mice::complete()}
\item
  on the 5 completed datasets \{\(y_{obs}, y_{imp, m}\)\}: 1) estimate
  Qhats for mu and sigma --\textgreater{} pool as mean()
  --\textgreater{} get Qbars of mu and sigma; 2) compute lambda as theta
  for conv diag; 3) perform lm to get Qhats for betas and r\^{}2
  --\textgreater{} use Qhats of beta as theta to compute conv diag
  --\textgreater{} pool Qhats of beta and r\^{}2 to get Qbars
\end{itemize}

@@

For each run of the imputation algorithm, we compute non-convergence
diagnostics \(\widehat{R}\) and \(AC\). We calculate \(\widehat{R}\) by
implementing Vehtari et al.'s \citeyearpar{veht19} recommendations, and
\(AC\) as the correlation between the \(t^{th}\) and the \((t+1)^{th}\)
iteration. The two methods are applied to four scalar summaries
\(\theta\): chain means (i.e., the mean in each \(y_{imp, m, j}\)),
chain variances (i.e., the variance in each \(y_{imp, m, j}\)), a
statistic of scientific interest \(\hat{Q}\) in each
\{\(y_{obs}, y_{imp, m}\)\}, and the first eigenvalue of the
variance-covariance matrix in each \{\(y_{obs}, y_{imp, m}\)\},
\(\lambda_1\).

The \(\widehat{R}\) and \(AC\) values with chain means as \(\theta\)s
are evaluated against the bias in univariate mean estimates. Similarly,
the performance measure for \(\widehat{R}\) and \(AC\) applied to chain
variances is the bias in estimated standard deviations. To evaluate the
performance of \(\widehat{R}\) and \(AC\) on the eigenvalues, we will
use bias in the coefficient of determination, \(R^2\). Additionally, we
will evaluate the estimated regression coefficients, the coverage rate
of the CI95\% of the regression estimates, and the CI length
(\textbf{see definitions in old Methods}). \textbf{remove the term
performance measure here and add an extra paragraph to define
performance measures as bias in all estimands, and coverage rate and CI
length of regression coefficients.}

\hypertarget{performance-measures.}{%
\subsubsection{Performance measures.}\label{performance-measures.}}

The impact of inducing non-convergence in the iterative imputation
algorithm is assessed by comparing \(\bar{Q}\)s with \(Q\)s. For each
\(Q\), we compute bias as \(\bar{Q} - Q\). For the regression
coefficients \(Q=\beta_{1,2,3}\), we also compute the empirical coverage
rate (CR). CR is defined as the percentage of simulation repetitions in
which the 95\% confidence interval (CI) around \(\bar{Q}\) covers the
true estimand \(Q\). Let
\[\text{CI} = \bar{Q} \pm t_{(M-1)} \times SE_{\bar{Q}},\] where
\(t_{(M-1)}\) is the quantile of a \(t\)-distribution with \(M-1\)
degrees of freedom, and \(SE_{\bar{Q}}\) is the square root of the
pooled variance estimate. Since under-estimating the variance of
\(\bar{Q}\) may yield spurious inferences, we also inspect CI width
(CIW): the difference between the lower and upper bound of the 95\%
confidence interval around \(\bar{Q}\).

The performance of two methods is evaluated against the bias and
confidence validity of the estimated \(Q\)s.

@@

\textbf{Add to this section: 1) emphasize that the plots are averages
across repetitions, not within MICE; 2) sample effects due to single
complete dataset.}

@@

\emph{Proper performance under a `missing completely at random'
missingness mechanism is necessary to demonstrate the appropriateness of
\(\widehat{R}\) and \(AC\) as non-convergence diagnostics. However,
results may not be extrapolated to other missingness mechanisms.}
